---
title: "Problem Set 4"
author: "Shafi Rubbani"
date: "2/24/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

#Install relevant libraries
library(tidyverse)
library(dplyr)
library(generics)
library(gt)
library(lubridate)
```

```{r read_data, include=FALSE}
#Read data into dataframe from file
polling_data <- read_csv(file = "ps_4_elections-poll-nc09-3.csv",
                #Select column formats
                 col_types =  cols(
                   .default = col_character(),
                   turnout_scale = col_double(),
                   turnout_score = col_double(),
                   w_LV = col_double(),
                   w_RV = col_double(),
                   #Final weight should be a double so we can perform arithmetic
                   final_weight = col_double(),
                   #Timestamp should be a datetime
                   timestamp = col_datetime(format = "")))
```

## Problem 1

```{r problem_1, include=FALSE}
#Filter data by "Dem" response, save results in a variable
dem_supporters <- polling_data %>% filter(response == "Dem") %>% count()

#Save results in a variable
rep_vs_und <- polling_data %>%
  #Filter data by "Rep" or "Und" response
  filter(response %in% c("Rep", "Und")) %>%
  #Group by response before summarizing
  group_by(response) %>% 
  #Count responses by party
  summarize(party_count = n()) %>% 
  #Summarize the difference between parties
  summarize(diff = party_count[1] - party_count[2])

#Save results in a variable
gender_labels <- polling_data %>%
  #Filter by difference between gender and gender_combined
  filter(gender != gender_combined) %>%
  #Count
  count()

#Save results in a variable
white <- polling_data %>%
  #Filter by race_eth "White" and file_race_black not "White"
  filter(race_eth == "White", file_race_black != "White") %>%
  #Count
  count()

#Save results in a variable
first_response_diff <- polling_data %>%
  #Filter by response "Rep" or "Dem"
  filter(response %in% c("Rep", "Dem")) %>%
  #Group by response
  group_by(response) %>%
  #Summarize earliest response by party
  summarize(first_response = min(timestamp)) %>%
  #Summarize difference between parties
  summarize(time_diff = round(first_response[2] - first_response[1], digits = 0))
```


There were `r dem_supporters` respondents who supported the Democratic candidate.

There were `r rep_vs_und` more respondents who favored the Republican candidate than who were Undecided.

There are two gender variables (gender and gender_combined). There are `r gender_labels` individuals for whom these variables have different values.

There are `r white` respondents listed as “White” under race_eth who are not listed as “White” under file_race_black.

The first response of Dem came `r first_response_diff` minutes (rounded to the nearest minute) before the first response of Rep.

## Problem 2

```{r problem_2, include=FALSE}
#Rename transformed data
response_chart <- polling_data %>% 
  #Filter out invalid responses
  filter(race_eth != "[DO NOT READ] Don't know/Refused") %>% 
  #Reorder races
  mutate(race_eth = fct_relevel(race_eth, c("White", "Black", "Hispanic", "Asian", "Other"))) %>%
  #Select relevant columns (party, race, and weight by response likelihood)
  select(response, race_eth, final_weight) %>% 
  #Group by race and response
  group_by(race_eth, response) %>% 
  #Sum weighted responses by group
  summarize(total = sum(final_weight)) %>%
  #Rearrange table to show party response proportions by race
  spread(key =  response, value = total)

#Set NA values to zero for arithemtic
response_chart[is.na(response_chart)] <- 0

response_chart <- response_chart %>%
  #Calculate total votes by race
  mutate(all = Dem + Rep + Und + `3`) %>% 
  #Calculate response proportions by race
  mutate(Dem = Dem / all) %>% 
  mutate(Rep = Rep / all) %>% 
  mutate(Und = Und / all) %>%
  #Hide third-party and total values
  select(-c(`3`, all)) %>% 
  ungroup() %>%
  #Set zero to NA again
  na_if(0)
```

```{r problem_2_table, echo=FALSE}
response_chart %>%
  #Make graphic table
  gt() %>% 
  #Set title
  tab_header(title = "Polling Results by Race in North Carolina 9th Congressional District") %>% 
  #Label columns
  cols_label(
    race_eth = "",
    Dem = "Democrat",
    Rep = "Republican",
    Und = "Undecided"
  ) %>%
  #Format proportions as percentages
  fmt_percent(columns = vars(Dem, Rep, Und),
              decimals = 0) %>%
  #Format NA values as dashes
  fmt_missing(columns = vars(Und), rows = 4)
```

## Problem 3

```{r problem_3}
polling_data %>% 
  #Filter out invalid responses
  filter(educ != "[DO NOT READ] Refused") %>%
  #Plot (re-ordered) education level against response weight distribution
  ggplot(aes(x = fct_rev(fct_relevel(educ, c("Graduate or Professional Degree", "Bachelors' degree", "Some college or trade school", "High school", "Grade school"))), y = final_weight)) + 
  #Flip axes
  coord_flip() +
  #Label title, subtitle, y axis, and caption
  labs(title = "More Educated Matter Less in North Carolina 9th",
       subtitle = "Poll gives more weight to people who are less likely to participate in polls",
       x = "",
       y = "Weight Given to Respondents in Calculating Poll Results",
       caption = "New York Times Upshot/Siena College 2018 live polls") +
  #Set aesthetic to violin plot
  geom_violin() +
  #Add visualization of datapoints
  geom_jitter(alpha = 0.3, size = 0.7, width = 0.2)
```

## Problem 4
```{r problem_4}
#I was interested in seeing the distribution of education level by gender.
education_by_gender <- polling_data %>%
  #Select relevant variables
  select(gender, educ) %>% 
  #Filter out irrelevant responses
  filter(educ != "[DO NOT READ] Refused") %>%
  #Group by gender and re-ordered education
  group_by(gender, educ = fct_rev(fct_relevel(educ, c("Graduate or Professional Degree", "Bachelors' degree", "Some college or trade school", "High school", "Grade school")))) %>% 
  #Count number per group
  summarize(count = n())
```

```{r problem_4_table}
education_by_gender %>% 
  #Plot gender against education counts
  ggplot(aes(x = gender, y = count, fill = educ)) +
  #Set aesthetic to stacked bar chart
  geom_col(position = 'fill') +
  labs(title = "Education of respondents by gender", x = "Gender", y = "Proportions")
```

The results are interesting. Men were more likely to have a high school or trade school level education, and women were more likely to have a bachelors' or professional degree. However, I am uncertain whether the differences in these proportions are statistically significant. I also note that the data only reflected the male and female genders and no others. This is likely an issue with data collection.